---
title: "Exploring"
format: html
editor: visual
---

```{r}
library(tidyverse)
library(readxl)
library(tidyverse)
library(ggplot2)
library(skimr)
library(corrplot)
library(GGally)
library(janitor)
```

```{r}
raw_data <- read_xlsx("data/dataset-uci.xlsx") %>% 
  clean_names() %>% 
  mutate(gallstone_status = factor(gallstone_status))

# 1. Quick structure and summary
summary(raw_data)
skim(raw_data)
glimpse(raw_data)
```

```{r}
# Convert target to factor
raw_data <- raw_data %>%
  clean_names() %>%
  mutate(gallstone_status = factor(gallstone_status))

# Confirm class balance
table(raw_data$gallstone_status)
```

What the EDA Suggests 🟢 1. Class Balance Very balanced dataset:

Gallstone = 1: 49.5% Gallstone = 0: 50.5%

✅ Great for training classification models — no resampling needed.

🟢 2. Boxplots and Correlations Your boxplots and pairwise correlations indicate some meaningful differences between gallstone vs. non-gallstone groups:

BMI, Total Body Fat Ratio, and Visceral Fat Area:

Slightly higher in the gallstone group.

Consistent with known risk factors (metabolic syndrome, obesity).

Glucose:

Appears right-skewed with some extreme outliers.

Higher glucose in gallstone group aligns with known diabetes/gallstone links.

HDL (Good cholesterol):

Positive correlation with gallstone presence is interesting — may warrant further stratification (e.g., by gender or BMI).

🔵 3. Correlation Matrix Strong internal correlations among body composition variables (e.g., TBFR, VFA, BMI).

Cholesterol and liver enzymes show lower correlations with gallstone status directly, but could still contribute in multivariate models.

```{r}


# 2. Missing values
colSums(is.na(raw_data))

# 3. Gallstone status distribution
raw_data %>%
  count(gallstone_status) %>%
  mutate(percent = round(n / sum(n) * 100, 1))

# 4. Histograms for all numeric variables
raw_data %>%
  select(where(is.numeric)) %>%
  pivot_longer(cols = everything(), names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  facet_wrap(~variable, scales = "free") +
  theme_minimal()

# 5. Correlation matrix
cor_data <- raw_data %>%
  select(where(is.numeric)) %>%
  cor(use = "complete.obs")

corrplot(cor_data, method = "color", tl.cex = 0.6, number.cex = 0.5)

# 6. Boxplots of selected variables by gallstone status
selected_vars <- c("age", "body_mass_index_bmi", "glucose",
                   "visceral_fat_area_vfa", "total_body_fat_ratio_tbfr_percent")

raw_data %>%
  select(gallstone_status, all_of(selected_vars)) %>%
  pivot_longer(cols = -gallstone_status, names_to = "feature", values_to = "value") %>%
  ggplot(aes(x = factor(gallstone_status), y = value, fill = factor(gallstone_status))) +
  geom_boxplot() +
  facet_wrap(~feature, scales = "free") +
  labs(x = "Gallstone Status", y = NULL, fill = "Gallstone") +
  theme_minimal()

# 7. Pairwise scatterplots for top variables
ggpairs(
  raw_data,
  columns = c("age", "body_mass_index_bmi", "glucose", "low_density_lipoprotein_ldl",
              "high_density_lipoprotein_hdl", "gallstone_status"),
  aes(color = factor(gallstone_status), alpha = 0.6)
)



```

🧠 So What's Left to Explore? 🟡 1. Model Interpretability & Explainability The paper focused on performance metrics but not explainability.

You can use:

SHAP (Shapley Additive Explanations) to visualize per-patient risk contributions.

LIME for local interpretability of models like GB or RF.

These help bridge ML models and clinical trust.

🟡 2. Subgroup or Stratified Analysis The paper gave aggregate results. You could explore:

Sex-specific predictors (e.g., does CRP matter more in men vs. women?).

Age groups or obesity classes as stratifiers.

Interaction effects: CRP × BMI or TBW × gender.

🟡 3. Alternative Feature Selection Methods They used ANOVA F-score.

You could test:

Recursive Feature Elimination (RFE)

Boruta for random forest-based selection.

LASSO (for sparse logistic regression).

Compare performance vs. ANOVA features.

🟡 4. Reproducibility & Transparency Rebuild their best models in R or Python and publish:

Open code + documentation (especially useful for clinical researchers).

Compare performance using different resampling strategies (e.g., repeated k-fold vs. 70/30 split).

🟡 5. External Validation or Transfer Learning As they mention, results are limited to their population.

If you can find an external gallstone dataset (e.g., from NHANES or another hospital):

Apply or fine-tune their models (domain adaptation).

Check generalizability.

🟡 6. Clinical Tool or Dashboard Package their best-performing model into an app or web tool.

Take patient inputs (CRP, vitamin D, TBW, etc.) and give risk prediction.

You can build a shiny dashboard (R) or Flask app (Python).

✨ Original Ideas You Could Publish or Share Topic Idea Model Explainability SHAP summary & force plots for key features Clinical Insight Subgroup analysis: CRP and TBW by gender or obesity level Reproducibility Open-source pipeline (data cleaning, modeling, evaluation) Feature Selection Comparison of ANOVA vs. Boruta vs. LASSO Visualization Interactive feature dashboard or model comparison tool Application Convert gradient boosting model into a clinical web tool

Let me know if you'd like to:

Rebuild one of their top models in R (e.g., xgboost with caret/tidymodels)?

Apply SHAP or LIME for interpretability?

Build a dashboard or app?

Run subgroup analysis?
